Learning to communicate using recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behavior, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each object.\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PI+WL, and combines pragmatic
  inference with word learning.}
\end{table}

@@

% XX note the relation to BGL model; they need uncertainty in the
% middle of the pragmatics specifically, which is where we end up
% putting it for other reasons. And in Bayesian framework uncertainty
% and learning must always coincide. Also the intuition that this is a
% general property of uncertainty for soft-max agents -- if you don't
% know where you are, good worlds have more peaked utility
% distributions than bad worlds, and thus more consistent behavior by
% other agents, so on average when you're guessing what other agents
% will do, good worlds contribute more to your average than bad
% worlds.

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behavior. Furthermore, even if only one partner is producing
the novel expressions, their behavior still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behavior should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (XX
% krauss 1966, branigan catchpole pickering 2011).

@@

% XXXX what can we say about the role pragmatics plays here? It's
% annoyingly non-trivial to sample from the equivalent model where the
% speaker just samples from a multinomial distribution and the
% listener inverts it. The listener's data is samples from a DM so
% they have conjugacy. But the speaker's data is listener
% interpretations, which necessarily do one tiny pragmatic inference
% step from using Bayes rule, which breaks conjugacy...

>>>>

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behavior, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each possible object.\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PI+WL, and combines pragmatic
  inference with word learning.}
\end{table}

@@

@@

Experimental results suggest that communicators are able to establish
novel, consensus-based communication systems. For example, adults
playing a communication game using only novel symbols with no
conventional meaning will typically converge on a set of new
conventions which allow them to accomplish their task
\cite{galantucci2005}. Or in a less extreme example, communicators
asked to refer to novel objects invent conventional names for them
over the course of repeated interactions (e.g., ``the ice skater'' for
an abstract figure vaguely resembling an ice skater,
\cite{clark1986}). From a pure learning perspective this behavior is anomalous,
however: Since both agents know perfectly well that there is no
underlying convention to discover, there is nothing to learn from the
other's behavior. Furthermore, even if only one partner is producing
the novel expressions, their behavior still becomes more regular
(conventional) over time, which would seem to rule out a role for
learning---even if there is some pattern in the expressions the
speaker chooses to use, there is certainly nothing for the
\textit{speaker} to learn by observing these patterns, and thus their
behavior should not change over time.
% An obvious hack might be to simply stipulate that speakers take
% their own utterances as training data for further learning, but this
% would fail to explain yet another wrinkle in this phenomenon:
% speakers only regularize their production over time if they have
% some kind of feedback indicating that the listener understands; if
% such feedback is removed, the phenomenon is dramatically reduced (X X
% krauss 1966, branigan catchpole pickering 2011).

@@

% X X X X what can we say about the role pragmatics plays here? It's
% annoyingly non-trivial to sample from the equivalent model where the
% speaker just samples from a multinomial distribution and the
% listener inverts it. The listener's data is samples from a DM so
% they have conjugacy. But the speaker's data is listener
% interpretations, which necessarily do one tiny pragmatic inference
% step from using Bayes rule, which breaks conjugacy...
